Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-18 paper/paper.tex
Refs

Termination and productivity are non-trivial properties of programs, and various
methods have been proposed for checking that these properties hold. A common
approach is guardedness checking~\cite{10.5555/646535.695850}, which is a
\emph{syntactic} check that definitions avoid the introduction of
non-normalizable terms. This sort of check generally looks for a
\emph{structural decrease} of arguments in the recursive calls of a function.
Coq uses such a check, and it works for many classic functional programming
patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
definitions will not be accepted by such checks.

In particular, the problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.

@@

This approach has been used to express nonstructurally recursive
algorithms in Agda: Copello et al.\ used sized types in a straightforward
formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.

@@

\bibliography{references.bib}
\appendix
\section{Polynomial Functors as Container Extensions}
\label{sec:pol-containers}
Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

>>>>

Termination and productivity are non-trivial properties of programs, and various
methods have been proposed for checking that these properties hold. A common
approach is guardedness checking~\cite{Eduardo94}, which is a
\emph{syntactic} check that definitions avoid the introduction of
non-normalizable terms. This sort of check generally looks for a
\emph{structural decrease} of arguments in the recursive calls of a function.
Coq uses such a check, and it works for many classic functional programming
patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
definitions will not be accepted by such checks.

In particular, the problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.

@@

This approach has been used to express nonstructurally recursive
algorithms in Agda: Copello et al.\ used sized types in a straightforward
formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
Coq as well, though it has not yet been adopted~\cite{ChanLB23}.

@@

\bibliography{references.bib}

\endinput
\appendix
\section{Polynomial Functors as Container Extensions}
\label{sec:pol-containers}
Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.
